
# Reproducible Research - Project course 1
---
author: "Manuel"
date: `r date()`
output: html_document
---

## Loading and preprocessing the data

```{r loading}
# As starting point I will create a directory ("data") where load the .zip file
if (!file.exists('data')){
    dir.create('data')
}

# And as next step I will download the information keeping record on when was downloaded
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL,destfile = 'data/data_project.zip')
unzip('data/data_project.zip',exdir = './data')
date_of_download <- date()

# Reading the input file and converting date column from factor to date
activity=read.csv("./data/activity.csv")
activity$date<-as.Date(activity$date)
```

Data has been loaded
    from `r fileURL`
    on `r date_of_download` 
    

## What is mean total number of steps taken per day?

```{r histogram & mean/median}

# Firstly I calculate the data I need to show in the histogram and required values
stepsday<-aggregate(steps ~ date, activity,sum)         # total steps/days
stepsmean<-aggregate(steps ~ date, activity,mean)       # mean steps/day
stepsmedian<-aggregate(steps ~ date, activity,median)   # median steps/day

library(ggplot2)
# I set centered the title of becoming plots
theme_update(plot.title = element_text(hjust = 0.5))
# Create the histogram of number of steps per day
ggplot(data=data.frame(stepsday), mapping = aes(x=date, y=steps)) + 
     geom_bar(stat = "identity", color="blue", fill="steelblue") +
     labs(x="Date")+
     labs(y="Steps")+
     ggtitle("Total number of steps taken per day")    

```

The mean and median of the total number of steps taken per day are:
    *mean*: **`r mean(stepsmean$steps)`**
    *median*: **`r median(stepsmedian$steps)`**

## What is the average daily activity pattern?

```{r average daily}

# get a data frame with the average of steps per interval
stepsIntervalmean<-aggregate(steps ~ interval, activity,mean)       

plot(stepsIntervalmean$interval, stepsIntervalmean$steps , 
     type="l",
     xlab="Interval",
     col="steelblue",
     ylab="Steps",
     main="Average steps per Interval")   

# I extract from the data frame the row which columne step has the max value
maxIntervalSteps<-stepsIntervalmean[stepsIntervalmean$steps ==
                      max(stepsIntervalmean$steps, 
                          na.rm = TRUE),]    
```

**`r maxIntervalSteps$interval`** is the *5-minute interval*, on average across all the days in the dataset, which contains the maximum number of steps

## Imputing missing values

```{r imputing missing values}

# Firstly I extract the total number of missing values in the dataset
missingValues=sum(is.na(activity))
```

The total number of *missing values* in the dataset is **`r missingValues`**

So in order to fill this missing value I use the following *strategy*: I will use the mean of the interval and will create a new dataframe that is equal to the original dataset but with the missing data filled in.

```{r new data set}

# Create a new data frame mergin the activity data frame and the data frame which
# contains step means by interval
newactivity=merge(activity, stepsIntervalmean, by="interval")

# Replace NA interval step values by the mean of steps  
newactivity$steps.x <- ifelse(is.na(newactivity$steps.x), 
                            newactivity$steps.y, 
                            newactivity$steps.x)

```

Now an histogram is created of the total number of steps taken each day and will calculate and report the mean and median total number of steps taken per day

```{r new histogram}
library(ggpubr)

newstepsday<-aggregate(steps.x ~ date, newactivity,sum)         # total steps/days
newstepsmean<-aggregate(steps.x ~ date, newactivity,mean)       # mean steps/day
newstepsmedian<-aggregate(steps.x ~ date, newactivity,median)   # median steps/day

hist1<-ggplot(data=data.frame(stepsday), mapping = aes(x=date, y=steps)) + 
    geom_bar(stat = "identity", color="blue", fill="steelblue") +
    labs(x="Date")+
    labs(y="Steps")+
    ggtitle("Total number of steps taken per day (with NAs)")  
hist2<-ggplot(data=data.frame(newstepsday), mapping = aes(x=date, y=steps.x)) + 
    geom_bar(stat = "identity", color="blue", fill="steelblue") +
    labs(x="Date")+
    labs(y="Steps")+
    ggtitle("Total number of steps taken per day (without NAs")
ggarrange(hist1, hist2,  
          ncol = 2, nrow = 1) 

```

The mean and median comparision of the total number of steps taken per day are:   
    *mean with NAs*: **`r mean(stepsmean$steps)`**  
    *mean without NAs*: **`r mean(newstepsmean$steps)`**    
    *median with NAs*: **`r median(stepsmedian$steps)`**     
    *median without NAs*: **`r median(newstepsmedian$steps)`**

## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays}
#First I create a column with the day of the date
activity=cbind(weekday=weekdays(activity$date),activity)

# Identify and replace weekday name by "weekday" or "weekend" values
activity$weekday <- ifelse(activity$weekday %in% c("sábado", "domingo"), 
                              "weekend", 
                              "weekday")

stepsIntervalWeekday<-aggregate(steps ~ interval+weekday, activity,mean)       

library(lattice)
xyplot(steps ~ interval | weekday, 
       data = stepsIntervalWeekday,
       type="l",
       layout = c(1, 2),
       main="Average across all weekday days or weekend",
       xlab = "Interval",
       ylab = "Average steps")
```

